Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-20 paper/analysis.tex
re-explain Chebyshev approximation

At this point, we acknowledge the existence of the XNysTrace estimator from \cite{epperly-2024-xtrace-making} which oftentimes outperforms the Nyström++ estimator. Moreover, a straight-forward extension of the XNysTrace estimator to the parameter-dependent setting seems to be in reach. However, what distinguishes \refalg{alg:nystrom-chebyshev-pp} for estimating spectral densities are two key observations --- using the cyclic invariance of the trace and the affine linear form of Chebyshev expansions (see \refsec{subsubsec:chebyshev-nystrom-implementation} for details) --- which it exploits to significantly speed up the computation. Unfortunately, we do not see a way of marrying these observations with the efficient implementation of XNysTrace described in \cite[Section 2.2]{epperly-2024-xtrace-making}, which limits its suitability for spectral density estimation.
%We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.

@@

\refthm{thm:hutchinson} shows that the Girard-Hutchinson estimator achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries. The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ queries (that is, matrix-vector products with random vectors), \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$. This parallels existing results for the constant case~\cite{meyer-2021-hutch-optimal,persson-2022-improved-variants}. In particular, Theorem 3.4 from~\cite{persson-2022-improved-variants} shows that $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ holds with high probability for (constant) SPSD $\mtx{B}$ when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$ in the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ from~\refequ{equ:nystrompp-trace-estimator}.

@@

%\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
\begin{proof}
For fixed $t$, consider the spectral decomposition $\mtx{B}(t) = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ with  
$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\lambda_1 \ge \dots \ge \lambda_n \ge 0$. For $k < n_{\mtx{\Omega}}$ (chosen arbitrarily at the moment), partition
    \begin{equation*}
        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
        \mtx{U} = \begin{bmatrix}
            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
        \end{bmatrix}
        \quad \text{and} \quad
        \mtx{\Lambda} =
        \begin{bmatrix}
            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
        \end{bmatrix},
    \end{equation*}
    we set
    $\mtx{\Omega}_1 := \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 := \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are independent Gaussian random matrices.
Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$, see also proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms}, yields the bound
    \begin{equation}
        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
        \label{equ:nystrom-proof-persson-bonud}
    \end{equation}
    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm. Following the proof of~\cite[Lemma 3]{meyer-2021-hutch-optimal}, we replace the the first term by a simpler bound:
    \begin{equation}
        \lVert \mtx{\Lambda}_2 \rVert _F
        \leq \sqrt{ \lambda_{k+1} (  \lambda_{k+1} + \cdots + \lambda_{n})}
        \leq \sqrt{ \Trace(\mtx{B}) / k \cdot \Trace(\mtx{B})}
        \leq \Trace(\mtx{B}) / \sqrt{k}.
        \label{equ:nystrom-proof-frobenius-trace}
    \end{equation}

>>>>

At this point, we acknowledge the existence of the XNysTrace estimator from \cite{epperly-2024-xtrace-making} which oftentimes outperforms the Nyström++ estimator. Moreover, a straight-forward extension of the XNysTrace estimator to the parameter-dependent setting seems to be in reach. However, what distinguishes \refalg{alg:nystrom-chebyshev-pp} for estimating spectral densities are two key observations --- using the cyclic invariance of the trace and the affine linear form of Chebyshev approximations (see \refsec{subsubsec:chebyshev-nystrom-implementation} for details) --- which it exploits to significantly speed up the computation. Unfortunately, we do not see a way of marrying these observations with the efficient implementation of XNysTrace described in \cite[Section 2.2]{epperly-2024-xtrace-making}, which limits its suitability for spectral density estimation.
%We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.

@@

\refthm{thm:hutchinson} shows that the Girard-Hutchinson estimator achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries. The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ queries (that is, matrix-vector products with random vectors), \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$. This parallels existing results for the constant case~\cite{meyer-2021-hutch-optimal,persson-2022-improved-variants}. In particular, \cite[Theorem 3.4]{persson-2022-improved-variants} shows that $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ holds with high probability for (constant) SPSD $\mtx{B}$ when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$ in the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ from~\refequ{equ:nystrompp-trace-estimator}.

@@

%\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
\begin{proof}
For fixed $t$, consider the spectral decomposition $\mtx{B}(t) = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ with  
$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\lambda_1 \ge \dots \ge \lambda_n \ge 0$. For $k < n_{\mtx{\Omega}}$ (chosen arbitrarily at the moment), partition
    \begin{equation*}
        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
        \mtx{U} = \begin{bmatrix}
            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
        \end{bmatrix}
        \quad \text{and} \quad
        \mtx{\Lambda} =
        \begin{bmatrix}
            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
        \end{bmatrix},
    \end{equation*}
    we set
    $\mtx{\Omega}_1 := \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 := \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are independent Gaussian random matrices.
Applying \cite[Theorem B.1]{persson-2023-randomized-lowrank} for $f(x) = x$, see also proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms}, yields the bound
    \begin{equation}
        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
        \label{equ:nystrom-proof-persson-bonud}
    \end{equation}
    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm. Following the proof of~\cite[Lemma 3]{meyer-2021-hutch-optimal}, we replace the the first term by a simpler bound:
    \begin{equation}
        \lVert \mtx{\Lambda}_2 \rVert _F
        \leq \sqrt{ \lambda_{k+1} (  \lambda_{k+1} + \cdots + \lambda_{n})}
        \leq \sqrt{ \Trace(\mtx{B}) / k \cdot \Trace(\mtx{B})}
        \leq \Trace(\mtx{B}) / \sqrt{k}.
        \label{equ:nystrom-proof-frobenius-trace}
    \end{equation}

### 2025-02-20 paper/intro.tex
re-explain Chebyshev approximation

\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for expanding matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

>>>>

\paragraph{New contributions.} We analyze three well-established randomized trace estimators when they are applied to parameter-dependent matrices: The Girard-Hutchinson estimator \cite{girard-1989-fast-montecarlo, hutchinson-1990-stochastic-estimator}, the trace of the Nyström low-rank approximation \cite{gittens-2013-revisiting-nystrom}, and the Nyström++ estimator \cite{persson-2022-improved-variants}. Interestingly, we can reuse the same randomization for computing the estimate at each value of the parameter $t$, which makes the estimators scale favorably with the number of parameter evaluations. We propose certain modifications to the methods from \cite{lin-2017-randomized-estimation} for approximating the smoothed spectral density $\phi_{\sigma}$ \refequ{equ:smoothed-spectral-density}. In particular, we use a more rigorous approach for approximating matrix functions in terms of Chebyshev polynomials and unify all of their methods in a single algorithm, with the desirable side effect of speeding up their \enquote{spectrum sweeping} method by orders of magnitude. Based on the analysis of the randomized trace estimators for parameter-dependent matrices, we provide bounds on the approximation error for our method.

### 2025-02-20 paper/numresults.tex
re-explain Chebyshev approximation

\color{black}
\textcolor{red}{
To do in this section:
\begin{itemize}
    \item Specify implementation details (no non-negative Chebyshev expansion, speed-up tricks, thresholding, ...)
    \item Clarify where code come from (faithful implementation, own code, ...)
    \item Give experimental setup (software version, hardware, single-threaded, ...)
\end{itemize}
}

@@

While theoretically an important tool, we waive using the non-negative Chebyshev expansion, since the slight indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-expansion} does not make a difference in finite precision arithmetic. We approximate the integrals for computing the $L^1$-errors in \refsec{sec:analysis} using a midpoint quadrature with $n_t \in \mathbb{N}$ nodes. For a fixed total number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$, we analyze the convergence behavior of the Chebyshev-Nyström++ method when run on the $1000 \times 1000$ matrix resulting from the finite difference discretization of the Hamiltonian for $n_c = 1$ described above (see \reffig{fig:convergence}).

@@

\begin{figure}[ht]
    \centering
    \input{plots/distribution.pgf}
    \caption{The Chebyshev-Nyström++ method for different ways of allocations a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter. We make the approximation error made in the Chebyshev expansion negligible by rescaling $m=16 / \sigma$ (based on \refthm{thm:chebyshev-error}).}
    \label{fig:distribution}
\end{figure}

>>>>

\color{black}
\textcolor{red}{
To do in this section:
\begin{itemize}
    \item Specify implementation details (no non-negative Chebyshev approximation, speed-up tricks, thresholding, ...)
    \item Clarify where code come from (faithful implementation, own code, ...)
    \item Give experimental setup (software version, hardware, single-threaded, ...)
\end{itemize}
}

@@

While theoretically an important tool, we waive using the non-negative Chebyshev approximation, since the slight indefiniteness of $g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ \refequ{equ:matrix-approximation} does not make a difference in finite precision arithmetic. We approximate the integrals for computing the $L^1$-errors in \refsec{sec:analysis} using a midpoint quadrature with $n_t \in \mathbb{N}$ nodes. For a fixed total number of random vectors $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}$, we analyze the convergence behavior of the Chebyshev-Nyström++ method when run on the $1000 \times 1000$ matrix resulting from the finite difference discretization of the Hamiltonian for $n_c = 1$ described above (see \reffig{fig:convergence}).

@@

\begin{figure}[ht]
    \centering
    \input{plots/distribution.pgf}
    \caption{The Chebyshev-Nyström++ method for different ways of allocations a total of $n_{\mtx{\Psi}} + n_{\mtx{\Omega}}=80$ random vectors to either the Nystr\"om low-rank approximation or the Girard-Hutchinson trace estimation for the Gaussian smoothing kernel with multiple different values of the smoothing parameter. We make the approximation error made in the Chebyshev approximation negligible by rescaling $m=16 / \sigma$ (based on \reflem{lem:non-negative-chebyshev-error}).}
    \label{fig:distribution}
\end{figure}
